Severe Weather Events Impacting Health and Economy
=======================================================

## Synopsis

This report answers two basic questions about severe weather events:  

1. Which storm events are the most harmful to population health, and 
2. Which storm events are most impactful to the economy.

To answer these questions, the storm database from the
U.S. National Oceanic and Atmospheric Administration's (NOAA) is utilized.  This database
tracks characteristics of major storms and weather events in the United States over time,
including estimates of any fatalities, injuries, and property and crop damage.

From this data, it's clear that tornadoes are by far most dangerous to people, while flooding are the most costly event types to the economy.  


## Data Processing
In order to prepare for the analysis, a few settings, packages and libraries must be loaded into the R programming environment:

```{r}
library(knitr)
library(Hmisc)
library(reshape)
library(ggplot2)
library(car)
```

### Getting and Loading the Data

From the Coursera "Reproducible Research" course
[file repository](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), we obtain the storm data in bzip archive format and extract it.

```{r cache=T,eval=F}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="repdata-data-StormData.csv.bz2")

bunzip2("repdata-data-StormData.csv.bz2", overwrite=T, remove=F)
```

The next step is to read the extracted file (in CSV format) into a data frame, which is then converted to a data.table for subsequent analysis.

```{r cache=T}
stormdata = read.csv("repdata-data-StormData.csv")
stormdata = data.table(stormdata)
```

Several of the columns within the dataset are cryptic when evaluated on its own.  The (National Climatic Data Center's
[record layout document](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc) describes the meaning of the columns and their values, but for simplicity sake in this analysis, a few key columns will be proper-cased and decodified, including the dollar figures and unit conversions.

Additionally, damages will be aggregated for fatalities and injuries for each event type by population and economic perspectives so that it can be plotted in the next section.

```{r cache=T}
stormdata$EVTYPE <- capitalize(tolower(stormdata$EVTYPE))

damages<-aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE , stormdata, sum)
dam<-melt(head(damages[order(-damages$FATALITIES,-damages$INJURIES),],10))

stormdata$PROPDMG<-stormdata$PROPDMG * as.numeric(Recode(stormdata$PROPDMGEXP, 
                                                         "'0'=1;
                                                         '1'=10;
                                                         '2'=100;
                                                         '3'=1000;
                                                         '4'=10000;
                                                         '5'=100000;
                                                         '6'=1000000;
                                                         '7'=10000000;
                                                         '8'=100000000;
                                                         'B'=1000000000;
                                                         'h'=100;
                                                         'H'=100;
                                                         'K'=1000;
                                                         'm'=1000000;
                                                         'M'=1000000;
                                                         '-'=0;
                                                         '?'=0;
                                                         '+'=0",
                                                         as.factor.result=FALSE))
stormdata$CROPDMG<-stormdata$CROPDMG * as.numeric(Recode(stormdata$CROPDMGEXP, 
                                                         "'0'=1;
                                                         '2'=100;
                                                         'B'=1000000000;
                                                         'k'=1000;
                                                         'K'=1000;
                                                         'm'=1000000;
                                                         'M'=1000000;
                                                         ''=0;
                                                         '?'=0",
                                                         as.factor.result=FALSE))

economic<-aggregate(cbind(PROPDMG, CROPDMG) ~ EVTYPE , stormdata, sum)
eco<-melt(head(economic[order(-economic$PROPDMG,-economic$CROPDMG),],10))
```

## Results ##
When analyzing the data, the total numbers are reviewed for each event in aggregate.
Yearly breakdown analysis is excluded in this basic analysis because there are many significant events that happen only once every few years.

### Human Casualties ###

* Question: Across the United States, which types of events are most harmful with respect to population health?

The below graph presents a barchart of the fatal and non-fatal Injuries, by event type.

```{r}
ggplot(dam, 
       aes(x=EVTYPE,
           y=value,
           fill=variable)) + 
        geom_bar(stat = "identity") + 
        coord_flip() +
        ggtitle("Harmful events") + 
        labs(x = "", 
             y="population affected") +
        scale_fill_manual (values=c("red","black"), 
                           labels=c("Deaths","Injuries"))

```

The chart clearly shows that tornados cause the highest rates of injuries, and excessive heat causes the highest rates of death.


### Economic impact ###

* Question: Across the United States, which types of events have the greatest economic consequences?

We now look at total property and crop damage per event type, for the 50
most costly event types.

By using the ggplot2 library we present a combined flipped barplot graph of the property and corp damages, by event type.

```{r}
ggplot(eco, 
       aes(x=EVTYPE,
           y=value,
           fill=variable)) + 
        geom_bar(stat = "identity") + 
        coord_flip() +
        ggtitle("Economic consequences") + 
        labs(x = "", 
             y="cost of damages in dollars") +
        scale_fill_manual (values=c("red","black"), 
                           labels=c("Property Damage","Crop Damage"))
```

As with dangerous event types, the graph highlights that floods have the greatest property damage and storm surge. 
